Commutative \(\Nat\)-rational Series of Polynomial Growth

### 2025-09-25 parts/newintro.tex
fix: pandoc + new intro

%! TeX program = xelatex
%! lang = en-US
% MAIN MATTER
\section{Introduction}
\label{introduction:sec}

\AP Given a semiring $\mathbb{S}$, and a finite alphabet $\Sigma$, the class of
\intro{(noncommutative) $\mathbb{S}$-rational series} is defined as functions from
$\Sigma^*$ to $\mathbb{S}$ that are computed by \kl{$\mathbb{S}$-weighted
automata} \cite{BERE10}. This computational model is a generalization of the
classical notion of non-deterministic finite automata to the weighted setting,
where transitions are labeled with elements of $\mathbb{S}$. The semantics of
\kl{$\mathbb{S}$-weighted automata} on a given word $w$ is defined by the sum
over all accepting runs reading $w$, of the product of the weights of the
transitions taken along this run. In this paper, we are interested in the case
where $\mathbb{S}$ equals $\Nat$ or $\Rel$, hence, in $\Nat$-rational series
($\intro*\NRat$) and $\Rel$-rational series ($\intro*\ZRat$). It is clear that
$\NRat$ is a proper subclass of $\ZRat$, and a longstanding open problem is to
provide an algorithm that decides whether a given $\ZRat$ is in $\NRat$
\cite{KARH77}.

\begin{problem}
    \label{n-in-z-rat:problem}
    Input: A $\ZRat$ $f$. Output: Is $f$ in $\NRat$?
\end{problem}

\Cref{n-in-z-rat:problem} recently received attention in the context of
\kl{polyregular functions} ($\intro*\Poly$), a computational model that aims to
generalize the theory of regular languages to the setting of string-to-string
functions \cite{BOJA18}. In the case of regular languages, \emph{star-free
languages} form a robust subclass of regular languages described equivalently
in terms of first order logic \cite{MNPA71}, counter-free automata
\cite{MNPA71}, or aperiodic monoids \cite{SCHU65}. Analogously, there exists a
\emph{star-free} fragment of \kl{polyregular functions} called \kl{star-free
polyregular functions} ($\intro*\SF$) \cite{BOJA18}. One open question in this area is to
decide whether a given \kl{polyregular function} is \kl{star-free}.

@@

\AP In order to approach decision problems on \kl{polyregular functions},
restricting the output alphabet to a single letter has proven to be a fruitful
method \cite{DOUE21,DOUE22}. Because words over a unary alphabet are
canonically identified with natural numbers, unary output \kl{polyregular
functions} are often called \kl{$\Nat$-polyregular functions} ($\NPoly$), and
their \emph{star-free} counterpart \kl{star-free $\Nat$-polyregular functions}
($\NSF$). Coincidentally, \kl{polyregular functions} with unary output forms a
subclass of \kl{$\Nat$-rational series}, namely the class of
\kl{$\Nat$-rational series} of \emph{polynomial growth}, i.e. the output of the
function is bounded by a polynomial in the size of the input. In \cite{CDTL23},
the authors introduced the class of \kl{$\Rel$-polyregular functions}
($\ZPoly$) as a subclass of \kl{$\Rel$-rational series} that generalizes
\kl{$\Nat$-polyregular functions} by allowing negative outputs, and showed that
membership in the \emph{star-free} subclass $\ZSF$ inside $\ZPoly$ is decidable
\cite[Theorem V.8]{CDTL23}. Although this could not be immediately leveraged to
decide $\NSF$ inside $\NPoly$, it was conjectured that $\NPoly \cap \ZSF =
\NSF$ \cite[Conjecture 7.61]{DOUE23}. It was believed that understanding the
membership problem of $\NPoly$ inside $\ZPoly$, that is, a restricted version
of \cref{n-in-z-rat:problem}, would be a key step towards proving $\NPoly \cap
\ZSF = \NSF$, which itself would give hope in designing an algorithm for
\cref{sf-polyregular:problem}. We illustrate in
\cref{previously-known-inclusions:fig} the known inclusions and related open
problems between the discussed classes of functions.

@@

As an intermediate step, we provide a complete and decidable characterization
of polynomials in $\Rat[\vec{X}]$ that can be computed using $\NRat$ (resp.
$\ZRat)$. These characterizations uncover a fatal flaw in the proof of a former
characterization of such polynomials \cite[Theorem 3.3, page 4]{KARH77}. We
also prove that this previous results holds for polynomials with at most two
indeterminates (\cref{lem:correct-covered-2}), which may explain why
it was not detected earlier. Furthermore, these characterizations provide
effective descriptions of polynomials that can be expressed in $\ZRat$ as those
obtained using integer combinations of products of \emph{binomial coefficients}
(called \kl{integer binomial polynomials}, defined page \kpageref{integer
binomial polynomial}) and similarly for $\NRat$ by introducing the notion of
\kl{strongly natural binomial polynomials} (defined page \kpageref{strongly
natural binomial polynomial}), which we believe has its own interest. Finally,
these characterizations demonstrate that polynomials expressible by $\ZRat$
(resp. $\NRat$) are exactly those expressible by $\ZSF$ (resp. $\NSF$), that
is, polynomials are inherently \emph{star free} functions.

>>>>

%! TeX program = xelatex
%! lang = en-US
% MAIN MATTER
\section{Introduction}\label{introduction:sec}

\AP\ Given a semiring $\mathbb{S}$, and a finite alphabet $\Sigma$, the class
of \intro{(noncommutative) $\mathbb{S}$-rational series} is defined as
functions from $\Sigma^*$ to $\mathbb{S}$ that are computed by
\kl{$\mathbb{S}$-weighted automata}~\cite{BERE10}. This computational model is
a generalization of the classical notion of non-deterministic finite automata
to the weighted setting, where transitions are labeled with elements of
$\mathbb{S}$. The semantics of \kl{$\mathbb{S}$-weighted automata} on a given
word $w$ is defined by the sum over all accepting runs reading $w$, of the
product of the weights of the transitions taken along this run. When the
semiring $\mathbb{S}$ is the Boolean semiring $\mathbb{B}$, one recovers the
classical class of regular languages via non-deterministic finite automata.

\begin{example}\label{ab-ba-power-two:ex}
  The function $f \colon \set{a,b}^* \to \Nat$ that maps a word $w$ to
  $2^{|w|_a - |w|_b} + 2^{|w|_b - |w|_a}$ is a
  $\Rat$-rational series.
  It can be computed by the following $\Rat$-weighted automaton.
  \begin{center}
    \begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid,auto]
      % initial state q_0 edge from above it
      \node[state,initial above
      ] (q_0)   {$q_0$};
      \node[state,accepting] (q_1) [right=of q_0] {$q_1$};
      \node[state,accepting] (q_2) [left=of q_0] {$q_2$};
      \path[->]
        (q_0) edge [bend left] node {$a \mid 2$} (q_1)
              edge [bend left] node {$a \mid 0.5$} (q_2)
              edge [bend right] node [swap] {$b \mid 2$} (q_2)
              edge [bend right] node [swap] {$b \mid 0.5$} (q_1)
        (q_1) edge [loop above] node {$a \mid 2$} ()
              edge [loop below] node {$b \mid 0.5$} (q_0)
        (q_2) edge [loop above]  node {$b \mid 2$} ()
              edge [loop below] node {$a \mid 0.5$} ()
              ;
        \node[draw=none] (outq1) [right=1cm of q_1] {$1$};
        \node[draw=none] (outq2) [left=1cm of q_2] {$1$};
        \node[draw=none] (outq0) [below=1cm of q_0] {$1$};
        \path[->]
          (q_1) edge [] (outq1)
          (q_2) edge [] (outq2)
          (q_0) edge [] (outq0)
          ;
    \end{tikzpicture}
  \end{center}
\end{example}

\AP\ When the semiring $\mathbb{S}$ is a commutative field, for instance in the
case of the rational numbers $\Rat$, the class of \kl{$\mathbb{S}$-rational
series} is well understood, and enjoys many decidability properties by virtue
of its correspondence with \emph{linear representations}~\cite{BERE10}.
From this correspondence, one can for instance decide whether two
$\Rat$-rational series are equal, or prove that a 
$\Real$-rational series that has values in $\Rat$ is in fact a 
$\Rat$-rational series.
However, the situation is much less clear when $\mathbb{S}$ is merely a
commutative ring or worse, a commutative semiring: the corresponding theories
of \emph{modules} or \emph{semimodules} are much more involved than the one for
vector spaces, and many questions are open. For instance, the function $f$ from
\cref{ab-ba-power-two:ex} cannot be computed by a $\Nat$-weighted automaton,
even though it is a $\Rat$-rational series that only takes natural values. Even
more, we can prove that $f$ cannot be computed by any $\Rel$-weighted
automaton.

In this paper, we are interested in the case where $\mathbb{S}$ equals $\Nat$
(semiring) or $\Rel$ (ring), hence, in $\Nat$-rational series ($\intro*\NRat$)
and $\Rel$-rational series ($\intro*\ZRat$). It is clear that $\NRat$ is a
proper subclass of $\ZRat$, and a longstanding open problem is to provide an
algorithm that decides whether a given $\ZRat$ is in $\NRat$~\cite{KARH77}.

\begin{problem}\label{n-in-z-rat:problem}
    Input: A $\ZRat$ $f$. Output: Is $f$ in $\NRat$?
\end{problem}

A similar open problem has been studied in the context of rational weighted
automata, where it amounts to deciding whether a given $\Rat$-weighted
automaton is equivalent to a $\Rat_+$-weighted automaton. Let us briefly
explain why this problem is equivalent to \cref{n-in-z-rat:problem}. First,
given a $\Rat$-weighted automaton $A$, one can construct a $\Rel$-weighted
automaton $A'$ by multiplying all weights by the least common multiple of the
denominators of the weights of $A$. Semantically, $A'(w) = c^{|w|} A(w)$ for
some constant $c \in \Nat$. If $A'$ is equivalent to an $\Nat$-weighted
automaton $B$, then $A$ is $c^{-|w|} B(w)$, hence equivalent to a
$\Rat_+$-weighted automaton. For the converse implication, if $A$ is a
$\Rat$-weighted automaton equivalent to a $\Rat_+$-weighted automaton, then
$A'$ is also equivalent to some $\Rat_+$-weighted automaton $B'$, but all the
outputs of $B'$ are natural numbers, and because $\Rat_+$ is a Fatou extension
of $\Nat$,\footnote{rework this} one concludes that $B'$ is equivalent to an
$\Nat$-weighted automaton $B''$~\cite{BERE10}.

\Cref{n-in-z-rat:problem} recently received attention in the context of
\kl{polyregular functions} ($\intro*\Poly$), a computational model that aims to
generalize the theory of regular languages to the setting of string-to-string
functions~\cite{BOJA18}. In the case of regular languages, \emph{star-free
languages} form a robust subclass of regular languages described equivalently
in terms of first order logic \cite{MNPA71}, counter-free automata
\cite{MNPA71}, or aperiodic monoids \cite{SCHU65}. Analogously, there exists a
\emph{star-free} fragment of \kl{polyregular functions} called \kl{star-free
polyregular functions} ($\intro*\SF$) \cite{BOJA18}. One open question in this area is to
decide whether a given \kl{polyregular function} is \kl{star-free}.

@@

\AP In order to approach decision problems on \kl{polyregular functions},
restricting the output alphabet to a single letter has proven to be a fruitful
method~\cite{DOUE21,DOUE22}. Because words over a unary alphabet are
canonically identified with natural numbers, unary output \kl{polyregular
functions} are often called \kl{$\Nat$-polyregular functions} ($\NPoly$), and
their \emph{star-free} counterpart \kl{star-free $\Nat$-polyregular functions}
($\NSF$). Coincidentally, \kl{polyregular functions} with unary output forms a
subclass of \kl{$\Nat$-rational series}, namely the class of
\kl{$\Nat$-rational series} of \emph{polynomial growth}, i.e. the output of the
function is bounded by a polynomial in the size of the input 
\cite{SCHU62}.
In \cite{CDTL23},
the authors introduced the class of \kl{$\Rel$-polyregular functions}
($\ZPoly$) as a subclass of \kl{$\Rel$-rational series} that generalizes
\kl{$\Nat$-polyregular functions} by allowing negative outputs, and showed that
membership in the \emph{star-free} subclass $\ZSF$ inside $\ZPoly$ is decidable
\cite[Theorem V.8]{CDTL23}. Although this could not be immediately leveraged to
decide $\NSF$ inside $\NPoly$, it was conjectured that $\NPoly \cap \ZSF =
\NSF$ \cite[Conjecture 7.61]{DOUE23}. It was believed that understanding the
membership problem of $\NPoly$ inside $\ZPoly$, that is, a restricted version
of \cref{n-in-z-rat:problem}, would be a key step towards proving $\NPoly \cap
\ZSF = \NSF$, which itself would give hope in designing an algorithm for
\cref{sf-polyregular:problem}. We illustrate in
\cref{previously-known-inclusions:fig} the known inclusions and related open
problems between the discussed classes of functions.

@@

As an intermediate step, we provide a complete and decidable characterization
of polynomials in $\Rat[\vec{X}]$ that can be computed using $\NRat$ (resp.
$\ZRat)$. These characterizations uncover a fatal flaw in the proof of a former
characterization of such polynomials \cite[Theorem 3.3, page 4]{KARH77}. We
also prove that this previous results holds for polynomials with at most two
indeterminates (\cref{lem:correct-covered-2}), which may explain why it was not
detected earlier. Furthermore, these characterizations provide effective
descriptions of polynomials that can be expressed in $\ZRat$ as those obtained
using integer combinations of products of \emph{binomial coefficients} (called
\kl{integer binomial polynomials}, defined page \kpageref{integer binomial
polynomial}) and similarly for $\NRat$ by introducing the notion of
\kl{strongly natural binomial polynomials} (defined page \kpageref{strongly
natural binomial polynomial}), which we believe has its own interest. In turn,
these characterizations demonstrate that polynomials expressible by $\ZRat$
(resp. $\NRat$) are exactly those expressible by $\ZSF$ (resp. $\NSF$), that
is, polynomials are inherently \emph{star free} functions.

Finally, we show that non-commutative polyregular functions in $\NPoly$ (resp.
$\NSF$) have a canonical representation in terms of their \emph{residuals}, and
propose a definition of star-free $\Rel$-rational series and $\Nat$-rational
series, that we believe is of independent interest.

### 2025-09-25 pnrs.tex
fix: pandoc + new intro

% For the date, we first check if the user has provided a date,
% and otherwise use the git meta inforamtion (if available).
\date{2025-09-22 10:31:27
+0200\footnote{00d33fe682fe4f73d3af284af9d9a8d732af14d2 -- branch theoretics at git@github.com:AliaumeL/polynomial-n-rational-series.git}}

@@

% Now, we create the document itself.
\begin{document}
% Generate the title page
\maketitle
% Print the abstract
\begin{abstract}
    This paper studies which functions computed by \(\Rel\)-weighted
    automata can be realised by \(\Nat\)-weighted automata, under two
    extra assumptions: commutativity (the order of letters in the input
    does not matter) and polynomial growth (the output of the function
    is bounded by a polynomial in the size of the input). We leverage
    this effective characterization to decide whether a function
    computed by a commutative \(\Nat\)-weighted automaton of polynomial
    growth is star-free, a notion borrowed from the theory of regular
    languages that has been the subject of many investigations in the
    context of string-to-string functions
    \paragraph{Keywords:}
    Rational series, Weighted automata, Polyregular
function, Commutative function
\paragraph{Repository:} \repositoryUrl
\paragraph{conference} \hyperlink{https://doi.org/10.4230/LIPIcs.STACS.2025.67}{STACS'25} \cite{lopez:LIPIcs.STACS.2025.67}
\end{abstract}

>>>>

% For the date, we first check if the user has provided a date,
% and otherwise use the git meta inforamtion (if available).
\date{2025-09-22 11:34:34
+0200\footnote{6d8e3bf714a8e6f6efba8be209bb9ae529722a78 -- branch theoretics at git@github.com:AliaumeL/polynomial-n-rational-series.git}}

@@

% Now, we create the document itself.
\begin{document}
% Generate the title page
\maketitle
% Print the abstract
\begin{abstract}
    This paper studies which functions computed by \(\Rel\)-weighted
    automata can be realised by \(\Nat\)-weighted automata, under two
    extra assumptions: commutativity (the order of letters in the input
    does not matter) and polynomial growth (the output of the function
    is bounded by a polynomial in the size of the input). We leverage
    this effective characterization to decide whether a function
    computed by a commutative \(\Nat\)-weighted automaton of polynomial
    growth is star-free, a notion borrowed from the theory of regular
    languages that has been the subject of many investigations in the
    context of string-to-string functions
    \paragraph{Keywords:}
    Rational series, Weighted automata, Polyregular
function, Commutative function
\paragraph{Repository:} \repositoryUrl
\paragraph{conference} \href{https://doi.org/10.4230/LIPIcs.STACS.2025.67}{STACS'25} \cite{lopez:LIPIcs.STACS.2025.67}
\end{abstract}
